\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-10-23 mers-structure.tex
Address #22 somewhat.
Add discussion paragraph that's 50% "we will all die" and 50% "eh, it'll be fine". Enough to close issue?

\subsection*{Transmissibility differences between zoonoses and pandemics}
Severe acute respiratory syndrome (SARS) coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
Since MERS-CoV was also able to cause significant pathogenicity in the human host it was inevitable that parallels would be drawn between MERS-CoV and SARS-CoV at the time of MERS discovery in 2012.
Although we describe the epidemiology of MERS-CoV from sequence data, indications that MERS-CoV has poor capacity to spread human-to-human existed prior to any sequence data.
SARS-CoV swept through the world in a short period of time, but MERS cases trickled slowly and were restricted to the Arabian Peninsula or resulted in self-limiting outbreaks outside of the region, a pattern strongly indicative of repeat zoonotic spillover.
Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
Even though our results show that MERS-CoV does not appear to present a serious global threat, we would like to highlight that its epidemiology is nonetheless concerning.

MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.

>>>>

\subsection*{Transmissibility differences between zoonoses and pandemics}
Severe acute respiratory syndrome (SARS) coronavirus, a Betacoronavirus like MERS-CoV, caused a serious epidemic in humans in 2003, with over 8000 cases and nearly 800 deaths.
Since MERS-CoV was also able to cause significant pathogenicity in the human host it was inevitable that parallels would be drawn between MERS-CoV and SARS-CoV at the time of MERS discovery in 2012.
Although we describe the epidemiology of MERS-CoV from sequence data, indications that MERS-CoV has poor capacity to spread human-to-human existed prior to any sequence data.
SARS-CoV swept through the world in a short period of time, but MERS cases trickled slowly and were restricted to the Arabian Peninsula or resulted in self-limiting outbreaks outside of the region, a pattern strongly indicative of repeat zoonotic spillover.
Infectious disease surveillance and control measures remain limited, so much like the SARS epidemic in 2003 or the H1N1 pandemic in 2009, zoonotic pathogens with $R_{0}>1.0$ are probably going to be discovered after spreading beyond the original location of spillover.
Even though our results show that MERS-CoV does not appear to present an imminent global threat, we would like to highlight that its epidemiology is nonetheless concerning.

MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans provides the substrate for a more transmissible variant of MERS-CoV to possibly emerge.
Previous research \citep{antia_role_2003} suggests that zoonotic pathogens become disproportionately more likely to evolve towards $R_{0}>1.0$ if $R_{0}$ is close to $1.0$ in the new host to begin with, which leaves MERS-CoV in a precarious position: on one hand our estimate of $R_{0}$ at $\approx 0.7$ makes MERS-CoV a potential pandemic hazard, on the other nothing is known of fitness landscapes or adaptive potential of MERS-CoV.
We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.
